Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2021-01-28 manuscript/plos/review_2.tex
revision 2 submitted

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{How can biological networks boost GWAS? A case study on susceptibility to familial breast cancer} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
GENESIS study group\textsuperscript{\textpilcrow},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Genome-wide association studies (GWAS) explore the genetic causes of complex diseases. However, they ignore the biological context of the genetic variants and genes under study. To address this shortcoming, one can use biological networks, which model functional relationships, to search for functionally related susceptibility loci. Many such network methods exist, each arising from different mathematical frameworks, pre-processing steps, and assumptions about the susceptibility mechanism. Unsurprisingly, this results in disparate solutions. To explore how to exploit these heterogeneous approaches, we selected six network methods and applied them to GENESIS, a GWAS on French women with non-BRCA familial breast cancer. First, we verified that network methods recovered more interpretable genes than a standard GWAS. We addressed the heterogeneity of their solutions by studying their overlap, computing what we called the \emph{consensus}. The key gene in this consensus solution was \emph{COPS5}, a gene related to multiple cancer hallmarks. Another issue we observed was that network methods were unstable, selecting very different genes on different subsamples of GENESIS. Therefore, we proposed a \emph{stable consensus} solution formed by the 68 genes most consistently selected across multiple subsamples. This solution was also enriched in genes associated with breast cancer susceptibility (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$). The most connected gene was \emph{CUL3}, a regulator of several genes linked to cancer progression. Lastly, we evaluated the biases of each method and the impact of their parameters on the outcome. In general, network methods preferred highly connected genes, even after random rewirings that stripped the connections of any biological meaning. In conclusion, we presented the advantages of network-guided GWAS, characterized their shortcomings, and provided strategies to address them. To compute the consensus networks, implementations of all six methods are available at \url{https://github.com/hclimente/gwas-tools}.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Genome-wide association studies (GWAS) scan thousands of genomes to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. However, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of \emph{core} genes are directly involved in the disease, but gene functions are deeply interrelated, and so many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. This study applies and compares six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS, thousands of unrelated samples are genotyped: the cases, suffering from the disease of interest, and the controls, taken from the general population. Then, a statistical test of association (e.g., based on logistic regression) is conducted between each SNP and the phenotype. Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment and deepening our understanding of the disease.

@@

Network-based discovery methods exploit the differential properties described above to identify disease genes on GWAS data \cite{azencott_network-guided_2016,huang_systematic_2018}. In essence, each gene receives a score of association with the disease, computed from the GWAS data, and a set of biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Multiple solutions have been proposed to this problem, often stemming from different mathematical frameworks and considerations of what the optimal solution looks like. For example, some methods restrict the problem to specific types of subnetworks. Such is the case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e., instances where both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, do not impose such strong constraints and search for subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy--Weinberg equilibrium in controls (P\=/value \textless~0.001), and those with genotyping data missing on more than 10\% of the samples. We also removed a subset of 20 duplicated SNPs in \emph{FGFR2}. We excluded the samples with more than 10\% missing genotypes. After controlling for relatedness, we excluded 17 additional samples (6 for sample identity error, 6 controls related to other samples, 2 cases related to an index case, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with lobular carcinoma \emph{in situ}, 1 with a \emph{BRCA1} or \emph{BRCA2} pathogenic variant detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls and 1\,280 cases, genotyped over 197\,083 SNPs.

@@

\paragraph{Gene network}
The mathematical formulations of the different network methods are compatible with any type of biological network (e.g., from protein interactions or gene co-expression). Here, we used protein-protein interaction networks (PPIN) for all network methods except SConES, as PPINs are interpretable, well-characterized, and the methods were designed to run efficiently on them. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless otherwise specified, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the PPIN. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we could map 142\,541 to a pair of gene symbols, involving 13\,619 genes. 12\,880 of those mapped to a genotyped SNP after quality control, involving 127\,604 interactions. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the PPIN and the nodes' scores. An insulated heat diffusion process captures the former: at initialization, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors and receives heat from its ``hotter'' neighbors while retaining part of its own (hence, \emph{insulated}). This process continues until a stationary state is reached, in which the temperature of the nodes does not change anymore, and results in a diffusion matrix $F$. $F$ is used to compute the similarity matrix $E$ that models exchanged heat as
\begin{eqnarray} 
E = F \operatorname{diag}(w(V)), 
\end{eqnarray}
where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. For any two nodes $i$ and $j$, $E_{ij}$ models the amount of heat that diffuses from node $j$ to node $i$. Hence, $E_{ij}$ can be interpreted as a (non-symmetric) similarity between those two nodes. To obtain densely connected solutions, HotNet2 prunes $E$, only preserving edges such that $w(E) > \delta$. Lastly, HotNet2 evaluates the statistical significance of the solutions by comparing their size to the size of PPINs obtained by permuting the node scores. We assigned the initial node scores as in Nakka et al. \cite{nakka_gene_2016}, giving a 0 to the genes with a VEGAS2 P-values of association with the disease, and -log\textsubscript{10}(P\=/value) to those likely to be. In the GENESIS dataset, the threshold separating both was a P\=/value of 0.125, which we obtained using a local FDR approach \cite{scheid_twilight;_2005}. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, which is robust to their values \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are parameters that control the sparsity and the connectivity of the model. The connectivity term penalizes disconnected solutions, with many edges between selected and unselected nodes. Given a $\lambda$ and an $\eta$, Eq~\ref{eq:scones} has a unique solution that SConES finds using a graph min-cut procedure. As in Azencott et al. \cite{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. In this case, the selected parameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI. We used the version on SConES implemented in the R package \texttt{martini} \cite{martini}.

@@

\subsubsection{Comparing SNP-methods to gene-methods and vice versa}
\label{methods:snp2gene}

@@

We evaluated multiple properties of the different methods (described in Sections~\ref{methods:algorithm_comparison} and \ref{methods:classifier}) through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original GENESIS dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the standard deviation of the different measures. Every method and repetition ran in the same computational settings (Section~\ref{methods:computation}).

@@

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries. BCAC included 13\,250\,641 SNPs genotyped or imputed on 228\,951 women of European ancestry, mostly from the general population \cite{Michailidou2017}. Through imputation, BCAC includes more SNPs than the iCOGS array used for GENESIS (Section~\ref{methods:data}). However, in all the comparisons in this paper we focused on the SNPs that passed quality control in GENESIS. Hence, we used the same Bonferroni threshold as in Section~\ref{methods:conventional} to determine the significant SNPs in BCAC. We also computed gene-scores in the BCAC data using VEGAS2, as in Section~\ref{methods:data}. In this case, we did use the summary statistics of all 13\,250\,641 available SNPs and the genotypes from European samples from the 1000 Genomes Project \cite{the_1000_genomes_project_consortium_global_2015} to compute the LD patterns. Since these genotypes did not include chromosome X, we excluded it from this analysis. All comparisons included only the genes in common between GENESIS and BCAC, so we used a different Bonferroni threshold ($1.66 \times 10^{-6}$) to call gene significance.

@@

Rewiring the PPIN while preserving the number of edges of each gene allowed to study the impact of the topology on the output of network methods. Indeed, the edges lose their biological meaning while the topology of the network is conserved. We produced 100 such rewirings by randomly swapping edges in the PPIN. We still scored the genes as described in Section~\ref{methods:methods}. We applied only four methods on the rewirings: heinz, dmGWAS, LEAN, and SigMod. We excluded HotNet2 and SConES since they took notably longer to run. As on the real PPIN, LEAN did not produce significant results on the rewirings either.

@@

Summary statistics for SNPs and genes are available at \url{https://github.com/hclimente/genewa}. We cannot share genotype data publicly for confidentiality reasons, but are available from GENESIS. Interested researchers can contact \href{mailto:nadine.andrieu@curie.fr}{nadine.andrieu(at)curie.fr}.

@@

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). At the SNP level, two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with the gene \emph{FGFR2}, the latter with \emph{CASC16} and the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. At the gene level, only \emph{FGFR2} was significantly associated with breast cancer (\nameref{sfig:snp_gene_manhattan}B).

@@

Despite these similarities, the consensus and the stable consensus included different genes. In the stable consensus network, the most central gene was \emph{CUL3}, which was absent from the previous consensus solution and had a low association score in both GENESIS and BCAC (P\=/values of 0.04 and 0.26, respectively). This gene was a component of Cullin-RING ubiquitin ligases. Encouragingly, it impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression was also linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the GWAS' ability to unravel the mechanisms leading to complex diseases has been called into question \cite{boyle_expanded_2017}. First, the omnigenic model proposes that gene functions are interwoven in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to discovering an uninformative wide-spread pleiotropy. Second, its conservative statistical framework hinders GWAS discovery. Network methods elegantly address these two issues by using both association scores and an interaction network to consider the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided discovery have been proposed in recent years. In this article we evaluated six of these methods (Section~\ref{methods:methods}) by applying them to the GENESIS study, a GWAS dataset on familial breast cancer (Section~\ref{methods:data}).

DmGWAS, Heinz, HotNet2, SConES, and SigMod all yielded compelling solutions, which include (but are not limited to) known breast cancer susceptibility genes (Section~\ref{results:separate_networks}). In general, the selected genes and SNPs were more central than most other genes and SNPs, agreeing with the observation that disease genes are more central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting any given random pair of nodes, making them more likely to be selected by network methods (Section~\ref{results:drawbacks}). However, we found support in the literature for the involvement of the selected highly central genes (\emph{COPS5}, \emph{FN1}, and \emph{CUL3}) in breast cancer susceptibility (Sections~\ref{results:consensus}, \ref{results:drawbacks}, and \ref{results:stable-consensus}). Despite these similarities, the methods' solutions were notably different. At one end of the spectrum, SConES and heinz preferred high scoring solutions, which were also small and hence did not shed much light on the disease's etiology. On the other end, dmGWAS, HotNet2 and SigMod gravitated towards lower scoring but larger solutions, which provided a wide overview of the biological context. While this deepened our understanding of breast cancer susceptibility and provided biological hypotheses, they required further analyses. For instance, we examined the centrality of the selected genes to understand how much that property was driving their selection (Section~\ref{results:drawbacks}). However, all solutions shared two drawbacks. First, they were all equally bad at discriminating cases from controls. However, the classification accuracy of network methods was similar to that of a classifier trained on the entire genome (Section~\ref{results:benchmark}), which suggests that cases and controls are difficult to separate in the GENESIS dataset. This may be due to limited statistical power, which reduces the ability to identify relevant SNPs. However, in any event, we do not expect to separate people who have or will develop cancer from others on the sole basis of their genomes, ignoring all environmental factors and chance events. Hence, network methods were preferable to the logistic regression classifier since they did ``no worse'' at classification while providing an interpretable solution. Second, all methods were remarkably unstable, yielding different solutions for slightly different inputs. This might partly have been caused by the instability of the P\=/values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P\=/values, which favored only the most extreme ones, led to improved stability. Another source of instability might have been the redundancy inherent to biological networks, a consequence of an evolutionary pressure to avoid single points of failure \cite{wagnerAlternativeRoutesMutational2007}. Hence, biological networks will often have multiple paths connecting two high-scoring nodes.

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Boosting GWAS using biological networks: a study on susceptibility to familial breast cancer} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
GENESIS study group\textsuperscript{\textpilcrow},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Genome-wide association studies (GWAS) explore the genetic causes of complex diseases. However, classical approaches ignore the biological context of the genetic variants and genes under study. To address this shortcoming, one can use biological networks, which model functional relationships, to search for functionally related susceptibility loci. Many such network methods exist, each arising from different mathematical frameworks, pre-processing steps, and assumptions about the network properties of the susceptibility mechanism. Unsurprisingly, this results in disparate solutions. To explore how to exploit these heterogeneous approaches, we selected six network methods and applied them to GENESIS, a nationwide French study on familial breast cancer. First, we verified that network methods recovered more interpretable results than a standard GWAS. We addressed the heterogeneity of their solutions by studying their overlap, computing what we called the \emph{consensus}. The key gene in this consensus solution was \emph{COPS5}, a gene related to multiple cancer hallmarks. Another issue we observed was that network methods were unstable, selecting very different genes on different subsamples of GENESIS. Therefore, we proposed a \emph{stable consensus} solution formed by the 68 genes most consistently selected across multiple subsamples. This solution was also enriched in genes known to be associated with breast cancer susceptibility (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$). The most connected gene was \emph{CUL3}, a regulator of several genes linked to cancer progression. Lastly, we evaluated the biases of each method and the impact of their parameters on the outcome. In general, network methods preferred highly connected genes, even after random rewirings that stripped the connections of any biological meaning. In conclusion, we present the advantages of network-guided GWAS, characterize their shortcomings, and provide strategies to address them. To compute the consensus networks, implementations of all six methods are available at \url{https://github.com/hclimente/gwas-tools}.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Genome-wide association studies (GWAS) scan thousands of genomes to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of cancers. However, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of \emph{core} genes are directly involved in the disease, but gene functions are deeply interrelated, and so many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. This study applies and compares six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS, thousands of unrelated samples are genotyped: the cases, suffering from the disease of interest, and the controls, taken from the general population. Then, a statistical test of association (e.g., based on logistic regression) is conducted between each SNP and the phenotype. Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in independent cohorts. Once the risk SNPs have been discovered, they can be used for risk assessment and deepening our understanding of the disease.

@@

Network-based discovery methods exploit the differential properties described above to identify disease genes using GWAS data \cite{azencott_network-guided_2016,huang_systematic_2018}. In essence, each gene receives a score of association with the disease, computed from the GWAS data, and a set of biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Multiple solutions have been proposed to this problem, often stemming from different mathematical frameworks and considerations of what the optimal solution looks like. For example, some methods restrict the problem to specific types of subnetworks. Such is the case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e., instances where both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, do not impose such strong constraints and search for subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

@@

We discarded SNPs with a minor allele frequency lower than 0.1\%, those not in Hardy--Weinberg equilibrium in controls (P\=/value \textless~0.001), and those with genotyping data missing on more than 10\% of the samples. We also removed a subset of 20 duplicated SNPs in \emph{FGFR2}. We excluded the samples with more than 10\% missing genotypes. After controlling for relatedness, we excluded 17 additional samples (6 for sample identity error, 6 controls related to other samples, 2 cases related to an index case, and 3 additional controls having a high relatedness score). Lastly, based on study selection criteria, 11 other samples were removed (1 control having cancer, 4 index cases with no affected sister, 3 half-sisters, 1 sister with lobular carcinoma \emph{in situ}, 1 with a \emph{BRCA1} or \emph{BRCA2} pathogenic variant detected in the family, 1 with unknown molecular diagnosis). The final dataset included 1\,271 controls, 1\,280 cases, and 197\,083 SNPs.

@@

\paragraph{Gene networks}
The mathematical formulations of the different network methods are compatible with any type of biological network (e.g., from protein interactions or gene co-expression). Here, we used protein-protein interaction networks (PPIN) for all gene-centric network methods, as PPINs are interpretable, well-characterized, and the methods were designed to run efficiently on them. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless otherwise specified, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the PPIN. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we could map 142\,541 to a pair of gene symbols, involving 13\,619 genes. 12\,880 of those mapped to a genotyped SNP after quality control, involving 127\,604 interactions. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

@@

\item[{HotNet2}] HotNet2 was developed to find connected subgraphs of genes frequently mutated in cancer \cite{leiserson_pan-cancer_2015}. To that end, it considers both the local topology of the PPIN and the nodes' scores. An insulated heat diffusion process captures the former: at initialization, the score of the node determines its initial heat; iteratively each node yields heat to its ``colder'' neighbors and receives heat from its ``hotter'' neighbors while retaining part of its own (hence, \emph{insulated}). This process continues until a stationary state is reached, in which the temperature of the nodes does not change anymore, and results in a diffusion matrix $F$. $F$ is used to compute the similarity matrix $E$ that models exchanged heat as
\begin{eqnarray} 
E = F \operatorname{diag}(w(V)), 
\end{eqnarray}
where $\operatorname{diag}(w(V))$ is a diagonal matrix with the node scores in its diagonal. For any two nodes $i$ and $j$, $E_{ij}$ models the amount of heat that diffuses from node $j$ to node $i$. Hence, $E_{ij}$ can be interpreted as a (non-symmetric) similarity between those two nodes. To obtain densely connected solutions, HotNet2 prunes $E$, only preserving edges such that $w(E) > \delta$. Lastly, HotNet2 evaluates the statistical significance of the solutions by comparing their size to the size of PPINs obtained by permuting the node scores. We assigned the initial node scores as in Nakka et al. \cite{nakka_gene_2016}, giving a 0 to the genes unlikely to be truly associated with the disease, and -log\textsubscript{10}(P\=/value) to those likely to be. In the GENESIS dataset, the threshold separating both was a P\=/value of 0.125, which we obtained using a local FDR approach \cite{scheid_twilight;_2005}. HotNet2 has two parameters: the restart probability \(\beta\), and the threshold heat \(\delta\). Both parameters are set automatically by the algorithm, which is robust to their values \cite{leiserson_pan-cancer_2015}. HotNet2 is implemented in Python \cite{hotnet2}.

@@

where \(\lambda\) and \(\eta\) are parameters that control the sparsity and the connectivity of the model. The connectivity term penalizes disconnected solutions, with many edges between selected and unselected nodes. Given a $\lambda$ and an $\eta$, Eq~\ref{eq:scones} has a unique solution that SConES finds using a graph min-cut procedure. As in Azencott et al. \cite{azencott_efficient_2013}, we selected \(\lambda\) and \(\eta\) by cross-validation, choosing the values that produce the most stable solution across folds. In this case, the selected parameters were \(\eta\)~=~3.51, \(\lambda\)~=~210.29 for SConES GS; \(\eta\)~=~3.51, \(\lambda\)~=~97.61 for SConES GM; and \(\eta\)~=~3.51, \(\lambda\)~=~45.31 for SConES GI. We used the version on SConES implemented in the R package \texttt{martini} \cite{martini,climente-gonzalezMartiniPackageGenomewide2021}.

@@

\subsubsection{Comparing SNP-based methods to gene-based methods and vice versa}
\label{methods:snp2gene}

@@

We evaluated multiple properties of the different methods (described in Sections~\ref{methods:algorithm_comparison} and \ref{methods:classifier}) through a 5-fold subsampling setting. We applied each method to 5 random subsets of the original GENESIS dataset containing 80\% of the samples (\emph{train set}). When pertinent, we evaluated the solution on the remaining 20\% (\emph{test set}). We used the 5 repetitions to estimate the average and the standard deviation of the different measures. Every method and repetition was run in the same computational settings (Section~\ref{methods:computation}).

@@

Additionally, we used the summary statistics from the Breast Cancer Association Consortium (BCAC), a meta-analysis of case-control studies conducted in multiple countries. BCAC included 13\,250\,641 SNPs genotyped or imputed on 228\,951 women of European ancestry, mostly from the general population \cite{Michailidou2017}. Through imputations, BCAC includes more SNPs than the iCOGS array used for GENESIS (Section~\ref{methods:data}). However, in all the comparisons in this paper we focused on the SNPs that passed quality control in GENESIS. Hence, we used the same Bonferroni threshold as in Section~\ref{methods:conventional} to determine the significant SNPs in BCAC. We also computed gene-scores in the BCAC data using VEGAS2, as in Section~\ref{methods:data}. In this case, we did use the summary statistics of all 13\,250\,641 available SNPs and the genotypes from European samples from the 1000 Genomes Project \cite{the_1000_genomes_project_consortium_global_2015} to compute the LD patterns. Since these genotypes did not include chromosome X, we excluded it from this analysis. All comparisons included only the genes in common between GENESIS and BCAC, so we used a different Bonferroni threshold ($1.66 \times 10^{-6}$) to call gene significance.

@@

Rewiring the PPIN while preserving the number of edges of each gene allowed to study the impact of the topology on the output of network methods. Indeed, the edges lose their biological meaning while the topology of the network is conserved. We produced 100 such rewirings by randomly swapping edges in the PPIN. We still scored the genes as described in Section~\ref{methods:methods}. We applied only four methods on the rewirings: heinz, dmGWAS, LEAN, and SigMod. We excluded HotNet2 and SConES since they took notably longer to run. As on the real PPIN, LEAN did not produce significant results on any of the rewirings either.

@@

Summary statistics for SNPs and genes are available at \url{https://github.com/hclimente/genewa}. We cannot share genotype data publicly for confidentiality reasons, but are available from GENESIS. Interested researchers can contact Séverine Eon-Marchais (severine.eon-marchais(at)curie.fr).

@@

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). At the SNP level, two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with the gene \emph{FGFR2}, the latter with \emph{CASC16} and the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, in studies on \emph{BRCA1} and \emph{BRCA2} carriers \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. At the gene level, only \emph{FGFR2} was significantly associated with breast cancer (\nameref{sfig:snp_gene_manhattan}B).

@@

Despite these similarities, the consensus and the stable consensus included different genes. In the stable consensus network, the most central gene was \emph{CUL3}, which was absent from the previous consensus solution and had a low association score in both GENESIS and BCAC (P\=/values of 0.04 and 0.26, respectively). This gene is a component of Cullin-RING ubiquitin ligases. It impacts the protein levels of multiple genes relevant for cancer progression \cite{Chen2016}, and its overexpression has also been linked to increased sensitivity to carcinogens \cite{Loignon2009}.

@@

In recent years, the GWAS' ability to unravel the mechanisms leading to complex diseases has been called into question \cite{boyle_expanded_2017}. First, the omnigenic model proposes that gene functions are interwoven in a dense co-function network. The practical consequence is that larger and larger GWAS will lead to discovering an uninformative wide-spread pleiotropy. Second, its conservative statistical framework hinders GWAS discovery. Network methods elegantly address these two issues by using both association scores and an interaction network to consider the biological context of each of the genes and SNPs. Based on what could be considered diverse interpretations of the omnigenic model, several methods for network-guided discovery have been proposed in recent years. In this article we evaluated six of these methods (Section~\ref{methods:methods}) by applying them to the GENESIS GWAS dataset on familial breast cancer (Section~\ref{methods:data}).

DmGWAS, Heinz, HotNet2, SConES, and SigMod all yielded compelling solutions, which include (but are not limited to) known breast cancer susceptibility genes (Section~\ref{results:separate_networks}). In general, the selected genes and SNPs were more central than most other genes and SNPs, agreeing with the observation that disease genes are more central \cite{pinero_uncovering_2016}. However, very central nodes are also more likely to be connecting any given random pair of nodes, making them more likely to be selected by network methods (Section~\ref{results:drawbacks}). However, we found support in the literature for the involvement of the selected highly central genes (\emph{COPS5}, \emph{FN1}, and \emph{CUL3}) in breast cancer susceptibility (Sections~\ref{results:consensus}, \ref{results:drawbacks}, and \ref{results:stable-consensus}). Despite these similarities, the methods' solutions were notably different. At one end of the spectrum, SConES and heinz preferred high scoring solutions, which were also small and hence did not shed much light on the disease's etiology. On the other end, dmGWAS, HotNet2 and SigMod gravitated towards lower scoring but larger solutions, which provided a wide overview of the biological context. While this deepens our understanding of breast cancer susceptibility and provides biological hypotheses, interpreting their solutions required further analyses. For instance, we examined the centrality of the selected genes to understand how much that property was driving their selection (Section~\ref{results:drawbacks}). However, all solutions shared two drawbacks. First, they were all equally bad at discriminating cases from controls. However, the classification accuracy of network methods was similar to that of a classifier trained on the entire genome (Section~\ref{results:benchmark}), which suggests that cases and controls are difficult to separate in the GENESIS dataset. This may be due to limited statistical power, which reduces the ability to identify relevant SNPs. However, in any event, we do not expect to separate people who have or will develop cancer from others on the sole basis of their genomes, ignoring all environmental factors and chance events. Hence, network methods were preferable to the logistic regression classifier since they did ``no worse'' at classification while providing an interpretable solution. Using an algorithm other than L1-penalized logistic regression might improve classification accuracy. Specifically, PRS-based strategies seem to perform slightly better on GENESIS \cite{lonjouGenePathwayLevel2021}. Second, all methods were remarkably unstable, yielding different solutions for slightly different inputs. This might partly have been caused by the instability of the P\=/values themselves in low statistical power settings \cite{halseyFickleValueGenerates2015}. Hence, heinz's conservative transformation of P\=/values, which favored only the most extreme ones, led to improved stability. Another source of instability might have been the redundancy inherent to biological networks, a consequence of an evolutionary pressure to avoid single points of failure \cite{wagnerAlternativeRoutesMutational2007}. Hence, biological networks will often have multiple paths connecting two high-scoring nodes.
